Mainstreaming Race Science

### 2023-09-04 paper/paper.Rmd
updating manuscript
updated manuscript to describe color-blind racism in more detail. added bonilla-silva's book to the bib file.

A key limitation of the current study stems from the inability of text mining techniques to draw on a broader cultural context than what is represented in the corpus.  Scholars of racism and US race relations have noted a shift in the way racist attitudes are understood and expressed, from *[something]* to "colorblind racism" *[Bonilla-Silva]*, defined as *[defn]*.  Colorblind racism involves the use of racialized "codes" or "dogwhistles," language that is superficially non-racial but carries racial implications, such as (in the US context) "violent inner city criminals" or "welfare queen" \cite{SaulDogwhistlesPoliticalManipulation2018}.  Alderfer notes that the racist implications of *The Bell Curve* only emerge gradually, with the opening of the book using the colorblind language of "group differences" \cite{AlderferScienceNonsciencePsychologists2003}.  Panofsky et al. show how the term "human biodiversity," originally developed for anti-racist purposes by biological anthropologist Jonathan Marks, has been inverted by white supremacists and is now a dogwhistle for biological racial hierarchy \cite{PanofskyHowWhiteNationalists2021}.  These cases suggest further that colorblindness can facilitate white supremacist appropriation of non-racial research — even contrary to the researchers' own intentions — as scientific racist readers re-interpret racial-neutral language as dogwhistles \cite{GillbornSoftlySoftlyGenetics2016, WillsAreClustersRaces2017}.

Techniques such as topic modeling may be able to account for discursive shifts from explicit to coded racial language, but only if the corpus contains a critical mass of documents using both sets of terms together.  This would seem to require either a relatively gradual transition, the inclusion of reflective commentary --- such as scholarship identifying and analyzing the transition --- or both.  It is unlikely that these requirements are met by the corpus used in this study.  So it is plausible that the topic models used here have some rate of "false negatives," documents in the corpus that engage in race science discourse but in a colorblind way that avoids the use of the explicit racial terms in the race-and-intelligence topic.

The phenomenon of colorblind racism is closely related to ambiguities that cannot be resolved even by human coders familiar with the broader cultural context.  Two relevant examples are the "Flynn effect," a secular increase in intelligence test scores noted by philosopher James Flynn; and studies of "national IQ," exemplified by the work of Richard Lynn.  The Flynn effect implies that environmental factors can create group differences in IQ that are comparable to Black-White differences, and thus directly challenges claims that the racial differences are biological.  Using methods widely regarded as ad hoc, cherry-picking, and generally unreliable, Lynn and collaborators have claimed that national IQ averages are correlated with national GDP.  Both the Flynn effect and "national IQ" are strongly associated with race-and-intelligence discourse in the US cultural context.  But scientific journal articles on these topics may not include any racial language at all.  For example, a meta-analysis of the Flynn effect does not discuss racial differences or use terms such as "White" or "Black" outside the reference list \cite{TrahanFlynnEffectMetaanalysis2014}.  Should such articles be coded as race science discourse?  We expect that even readers who accept our definitions of race science and race science discourse will reasonably disagree on how to answer this question.  Topic models, of course, are not even capable of representing this essential ambiguity.

>>>>

A key limitation of the current study stems from the inability of text mining techniques to draw on a broader cultural context than what is represented in the corpus.  Scholars of racism and US race relations have noted a shift in the way racist attitudes are understood and expressed, from overt and direct to a subtle "color-blind" racism \cite{bonillasilva2022racism}. Color-blind racism represents a change in the primary mechanisms by which white privilege is maintained, using non-racialized language in furtherance of the racial status quo both institutionally and individually, and increasingly reliant on the invisibility of socio-cultural mechanisms to preserve inequalities. Color-blind racism involves the use of racialized "codes" or "dogwhistles," language that is superficially non-racial but carries racial implications, such as (in the US context) "violent inner city criminals" or "welfare queen" \cite{SaulDogwhistlesPoliticalManipulation2018}.  Alderfer notes that the racist implications of *The Bell Curve* only emerge gradually, with the opening of the book using the color-blind language of "group differences" \cite{AlderferScienceNonsciencePsychologists2003}.  Panofsky et al. show how the term "human biodiversity," originally developed for anti-racist purposes by biological anthropologist Jonathan Marks, has been inverted by white supremacists and is now a dogwhistle for biological racial hierarchy \cite{PanofskyHowWhiteNationalists2021}.  These cases suggest further that color-blindness can facilitate white supremacist appropriation of non-racial research — even contrary to the researchers' own intentions — as scientific racist readers re-interpret racial-neutral language as dogwhistles \cite{GillbornSoftlySoftlyGenetics2016, WillsAreClustersRaces2017}.

Techniques such as topic modeling may be able to account for discursive shifts from explicit to coded racial language, but only if the corpus contains a critical mass of documents using both sets of terms together.  This would seem to require either a relatively gradual transition, the inclusion of reflective commentary --- such as scholarship identifying and analyzing the transition --- or both.  It is unlikely that these requirements are met by the corpus used in this study.  So it is plausible that the topic models used here have some rate of "false negatives," documents in the corpus that engage in race science discourse but in a color-blind way that avoids the use of the explicit racial terms in the race-and-intelligence topic.

The phenomenon of color-blind racism is closely related to ambiguities that cannot be resolved even by human coders familiar with the broader cultural context.  Two relevant examples are the "Flynn effect," a secular increase in intelligence test scores noted by philosopher James Flynn; and studies of "national IQ," exemplified by the work of Richard Lynn.  The Flynn effect implies that environmental factors can create group differences in IQ that are comparable to Black-White differences, and thus directly challenges claims that the racial differences are biological.  Using methods widely regarded as ad hoc, cherry-picking, and generally unreliable, Lynn and collaborators have claimed that national IQ averages are correlated with national GDP.  Both the Flynn effect and "national IQ" are strongly associated with race-and-intelligence discourse in the US cultural context.  But scientific journal articles on these topics may not include any racial language at all.  For example, a meta-analysis of the Flynn effect does not discuss racial differences or use terms such as "White" or "Black" outside the reference list \cite{TrahanFlynnEffectMetaanalysis2014}.  Should such articles be coded as race science discourse?  We expect that even readers who accept our definitions of race science and race science discourse will reasonably disagree on how to answer this question.  Topic models, of course, are not even capable of representing this essential ambiguity.
